---
title: "Event-Based Modeling of FXTAS Symptoms Progression"
output: 
  word_document: default
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{kde-ebm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>",
  fig.width = 4, 
  fig.height = 7,
  include = FALSE
)
```


# Methods

```{r setup}
devtools::load_all()
# library(rSuStaIn)
library(dplyr)
library(magrittr)
library(pander)
library(reticulate)

```

```{r biomarker list, include = FALSE}

tremors = c(
  "intention tremor" = 1,
  "resting tremor" = 1,
  "postural tremor" = 1,
  "intermittent tremor hx" = 1,
  "any tremor (excluding head)" = 1
)

categorical_biomarker_directions = c(
  
  "Cerebral Atrophy" = 1,
  "Cerebellar Atrophy" = 1,
  "Cerebral WM Hyperintensity" = 1,
  "Cerebellar WM Hyperintensity" = 1,
  "MCP-WM Hyperintensity" = 1)

continuous_biomarker_directions = 
  c(
    "ataxia severity exam" = 1,
    "MMSE total score" = -1,
    "BDS-2 Total Score" = -1,
    "SCL90: Somatization" = 1,
    "SCL90: Obsessive-Compulsive" = 1,
    "SCL90: Anxiety" = 1,
    "SCL90: Depression" = 1,
    "SCL90: Hostility" = 1,
    "SCL90: Phobia" = 1,
    "SCL90: Paranoid Ideation" = 1,
    "SCL90: Psychoticism" = 1,
    "SCL90: Global Severity Index" = 1,
    "SCL90: Interpersonal Sensitivity" = 1,
    "SCL90: Positive Symptom Distress Index" = 1,
    "SCL90: Positive Symptom Total" = 1,
    "Verbal: IQ Score" = -1,
    # "Verbal Comprehension: Composite Score (VCI)" = -1,
    "Perceptual Reasoning: Composite Score (PRI)" = -1,
    "Working Memory: Composite Score (WMI)" = -1,
    "Processing Speed: Composite Score (PSI)" = -1,
    "Full Scale: IQ Score" = -1,
    
    NULL
  )

continuous_vars_that_dont_work = 
  c(
    "Full Scale: Composite Score (FSIQ)" = -1, # not enough controls
    # cantab measures not available for controls:
    "SWM Between errors" = -1,
    "SST Median correct RT on GO trials" = -1,
    "RVP A signal detection" = -1,
    "OTS Problems solved on first choice" = -1,
    "PAL Total errors (adjusted)" = -1,
    "RTI Five-choice movement time" = -1,
    "FXTAS Stage (0-5)" = 1, # doesn't work, because there's no variation among controls
    NULL
  )

biomarker_directions = c(
  tremors,
  continuous_biomarker_directions
)

biomarkers = names(biomarker_directions)

```

```{R data processing}

cur_data = 
  visit1 |> 
  dplyr::select(
    all_of(biomarkers),
    FX,
    `FX*`,
    `FXS ID`,
    `Age at visit`
  ) |> 
  dplyr::filter(!is.na(FX))

### Mann-Whitney tests
# ```{r, "Mann-Whitney tests"}
# for (cur_biomarker in biomarkers)
# {
#   cat("### ", cur_biomarker)
#   wilcox.test(
#   formula = formulaic::create.formula(cur_biomarker, "FX", dat = cur_data)$formula, 
#   data = cur_data,
#   na.action = na.exclude) |> print()
# }

# ```{r, "prepare for python"}

X = 
  cur_data |> 
  dplyr::select(all_of(biomarkers)) |> 
  dplyr::mutate(
    across(where(is.factor), as.numeric)
  ) |> 
  as.matrix()

X_cases = 
  cur_data |> 
  dplyr::filter(FX) |> 
  dplyr::select(all_of(biomarkers)) |> 
  as.matrix()

y = cur_data |> dplyr::pull(FX) |> as.integer() |> as.matrix()

```


```{r "ggplot graphs"}

if(FALSE)
{
  library(ggplot2)
  for (cur in biomarkers)
  {
    
    temp = ggplot(data = cur_data, aes(x = get(cur), group = FX)) + 
      geom_histogram() + 
      facet_wrap(~FX, ncol = 1) +
      ggplot2::ggtitle(cur)
    
    print(cur)
    print(temp)
  }
  
}

```

## Notation

Following the Event-Based Model (EBM) analysis developed in Fonteijn et al 2012, Firth et al 2020, Oxtoby et al 2021, etc., we consider a set of $p$ biomarkers, $X_1$ ... $X_p$, measured on a cross-sectional case-control survey of $n$ patients and $m$ controls.

## Assumptions

We make the following assumptions:

1.  Biomarker measurements from different individuals are mutually independent.
2.  The distribution of each biomarker $X_j$ depends on a corresponding unobserved binary event, $E_j$.
3.  All patients are progressing along the same ordering sequence $\vec S\in \{1:p\}!$ of events $E_1$ - $E_p$, where $\{1:p\}!$ represents the set of all $p!$ orderings of the numbers $1:p$.
4.  Each patient is at a particular position $K\in 1:p$ within that sequence, such that $E_{S(j)}=1,\ j \leq K$ and $E_{S(j)}=0,\ j>K$.
5.  The distribution of $K$ given $\vec S$ is uniform; i.e., $\mathrm{p}(K=k|\vec S = \vec s)=1/p$, $k \in \{1:p\}$.
6.  The Bayesian prior distribution over the possible values of $\vec S$ is uniform; i.e., $\mathrm{p}(\vec S = \vec s)=1/p!,\ \vec s\in \{1:p\}!$.

We estimate $S$ using a two-step procedure:

## Analysis Step 1: Estimate biomarker distribution models conditional on corresponding events


```{python densities, include = FALSE}

from kde_ebm.mixture_model import fit_all_kde_models, fit_all_gmm_models, get_prob_mat

from kde_ebm.plotting import mixture_model_grid, mcmc_uncert_mat, mcmc_trace, stage_histogram

from kde_ebm.mcmc import mcmc, parallel_bootstrap, bootstrap_ebm, bootstrap_ebm_fixedMM, bootstrap_ebm_return_mixtures

X = r.X
y = r.y.flatten().astype(int)
e_disease_direction = r.biomarker_directions
e_labels = r.biomarkers


kde_mixtures = fit_all_kde_models(X, 
y,
implement_fixed_controls = True,
patholog_dirn_array      = e_disease_direction,
verbose = False)


```

In the first analysis phase, we analyze each biomarker separately. We model the distribution of each biomarker $X_j$ conditional on the corresponding event $E_j$; i.e., we model $\mathrm{p}(x_j|E_j=0)$ and $\mathrm{p}(x_j | E_j=1)$, using nonparametric kernel density estimation. To fit these models, we begin by labeling all controls as pre-event ($E_j=0$) and all cases as post-event ($E_j=1$), and then iteratively relabel some of the cases and controls, re-estimating $\mathrm{p}(x_j \mid E_j=1)$ and $\mathrm{p}(x_j \mid E_j=0)$ accordingly. [This step isn't given much detail in the published articles, and the code implementation is convoluted; I intend to do a deeper dive into the code and write out all the details.]

## Analysis Step 2: Estimate event sequence

```{python "MCMC sequencing", include = FALSE}

import numpy
numpy.random.seed(0)

mcmc_samples = mcmc(r.X_cases, 
kde_mixtures, 
n_iter = 10**5, 
greedy_n_init = 20,
plot = False)

```

In the second analysis phase, we compute the Bayesian posterior distribution of $\vec S$ given the observed data $\mathbf X$, i.e., $\mathrm{p}(\vec S \mid \mathbf X)=\mathrm{p}(\mathbf X \mid \vec S)\mathrm{p}(\vec  S)/\mathrm{p}(\mathbf X)$, using a Markov-Chain Monte Carlo (MCMC) algorithm. To do so, we compute $\mathrm{p}(\mathbf X \mid \vec S = \vec s)$ using the following decomposition:

$$
\mathrm{p}(\mathbf X|\vec S = \vec s) = \prod_{i=1}^n \mathrm{p}(\vec X_i | \vec S = \vec s)
$$

(using the assumption of independence between patients)

$$ 
\begin{align}
\mathrm{p}(\vec X_i \mid \vec S = \vec s) &= \sum_{k=0}^p \mathrm{p}(\vec X_i= \vec x_i \mid \vec S = \vec s,\ K_i=k) * \mathrm{p}(K_i = k \mid \vec S = \vec s)\\
&= \sum_{k=0}^p \mathrm{p}(\vec X_i= \vec x_i \mid \vec S = \vec s,\ K_i=k) * (1/p!)\\
&= (1/p!) * \sum_{k=0}^p \mathrm{p}(\vec X_i= \vec x_i \mid \vec S = \vec s,\ K_i=k)\\
&\propto \sum_{k=0}^p \mathrm{p}(\vec X_i= \vec x_i \mid \vec S = \vec s,\ K_i=k)
\end{align}
$$

(using the assumption that $\mathrm{p}(K_i = k \mid \vec S = \vec s)$ is uniformly distributed)

$$
\mathrm{p}(\vec X_i= \vec x_i \mid \vec S = \vec s,K_i=k) =
\left(\prod_{j=1}^k \hat{\mathrm{p}}(X_{ij}=x_{ij} | E_{s(j)}=1)\right)
\left(\prod_{j=k+1}^p \hat{\mathrm{p}}(X_{ij}=x_{ij} | E_{s(j)}=0)\right)
$$

(using the assumption that all patients follow the same sequence of events)

In the last expression, we plug in the estimated models $\hat{\mathrm{p}}(x_j|E_j=0)$ and $\hat{ \mathrm{p}}(x_j|E_j=1)$ from phase 1.

The MCMC sampling algorithm is as follows:

1.  Let $\vec s_t$ be the current sampled value of $\vec S$.
2.  Construct a new sequence $\vec s'$ by randomly permuting two elements of $s_t$.
3.  Calculate $a = \mathrm{p}(\mathbf X \mid \vec S = \vec s') / \mathrm{p}(\mathbf X \mid \vec S = \vec s_t)$
4.  Set $\vec s_{t+1} \leftarrow \vec s'$ with probability $\min(a, 1)$; otherwise set $\vec s_{t+1} \leftarrow \vec s_{t}$.

## Biomarkers analyzed

The analysis implemented in Oxtoby et al 2021 can only handle biomarkers which are measured as continuous variables and which vary for both cases and controls. So far, we have included the following variables:


```{r, include = TRUE}

tibble(biomarkers) |> pander()

```

# Results

## Biomarker distribution models

Here, we show the estimated distributions $\hat{\mathrm{p}}(X_j | E_j =1)$ and $\hat{\mathrm{p}}(X_j | E_j =0)$ for each biomarker (orange = patients, blue = controls, purple is the probability that the event has occurred given the biomarker value (i.e., $\mathrm{p}(E_j = 1 | X_j = x)$):

```{python, include = FALSE}
import matplotlib.pyplot as plt
mixture_model_grid(X,
y,
kde_mixtures,
score_names=r.biomarkers,
class_names=['Controls','Patients'],
max_cols = 2)

```

```{python "mixture model graphs", include = TRUE, fig.width = 4, fig.height = 9}

plt.show()

```

## Event sequence

Here, we show a heatmap of the estimated position of each biomarker event in the sequence $\vec S$:

```{python, include = TRUE, fig.width = 5}
# f,a = mcmc_uncert_mat(mcmc_samples, ml_order=None, score_names=e_labels)
# plt.show()

```

```{r "heatmap in r", include = TRUE, fig.width = 9}
heatmap_data = 
  py$mcmc_samples |> 
  magrittr::set_attr("biomarker labels", biomarkers)  |> 
  event_order_heatmap()

heatmap_data |> 
  plot_heatmap() |> 
  # magrittr::add(theme_ipsum()) |> 
  print() 
```

```{python}

from kde_ebm.staging import ebm_staging
#* Staging
#* Maximum-likelihood model stage: could include longitudinal data, including followups not used to train the EBM

[prob_mat_ml, 
stages_long_ml, 
stage_likelihoods_long_ml, 
stages_long_ml_expected] = ebm_staging(x=X,
mixtures=kde_mixtures,
samples=mcmc_samples
)

stages_long = stages_long_ml

```

Distribution of stages by FX status (CGG $\ge$ 55):

```{r, include = TRUE}

cur_data$stages_ML = py$stages_long_ml_expected
py$stage_likelihoods_long_ml[47,]
library(ggplot2)

plot1 = cur_data |> 
  ggplot(aes(x = stages_ML)) +
  facet_grid(~`FX*`) +
  geom_histogram() + 
  expand_limits(x = c(0, length(biomarkers)))

print(plot1)
```

```{r}
library(ggplot2)
cur_data |> 
  dplyr::filter(!FX) |> 
  dplyr::select(stages_ML, `FXS ID`, `Age at visit`) |> 
  arrange(stages_ML) |>  
  ggplot() +
  aes(
    y = .data$stages_ML,
    x = .data$`Age at visit`
  ) + 
  geom_point()
```



# Discussion

## Questions about Analysis

-   In Analysis Phase 2 (event sequencing) should we use the data from cases only or both cases and controls? Oxtoby's tutorial uses both, but that doesn't make much sense; controls aren't expected to follow the same event sequence.

-   Are the assumptions reasonable?

-   Is the mixture-modeling step (Analysis Phase 1) reasonable?

-   Should it be separate from the sequencing step (phase 2)?





