---
title: "parkinsonian-features"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rSuStaIn)
```

```{r}
library(tidyverse)
pfs = c("Masked faces",
        "Increased tone",
        "pill-rolling tremor exam",
        "Stiff gait")
parkinsonian_vars =
  c(
    "Parkinsons",
    "parkinsonian features",
    pfs)
parkinsonian_data = 
  trax_gp34_all |> dplyr::select(all_of(
    c(
      "Study",
      "FXS ID",
      "Event Name",
      "Visit Date",
      parkinsonian_vars))) |> 
  dplyr::mutate(
    any_pf =if_any(
      all_of(pfs),
      ~ .x == "Yes"
      )
  )
parkinsonian_data |> summary()
parkinsonian_data |> dplyr::filter(`parkinsonian features` == "Yes") |> 
  dplyr::select(-`parkinsonian features`) |> ftable(row.vars = 1:2)

parkinsonian_data |> dplyr::filter(`parkinsonian features` == "No")
```

Does anyone have `pk features` = Yes but no individual features? No:

```{r}
parkinsonian_data |> 
  dplyr::filter(`parkinsonian features` == "Yes", !any_pf)

```

However, there are cases where there are individual parkinsonian features but `parkinsonian features`  == "No":

```{r}

parkinsonian_data |> 
  dplyr::filter(`parkinsonian features` == "No", any_pf)

parkinsonian_data |> 
  select(`parkinsonian features`, any_pf) |> 
  table(useNA = 'always')
```


```{r}

parkinsonian_data |> 
  slice_head(n = 1, by = `FXS ID`) |> 
  dplyr::filter(
    `parkinsonian features` == "No", any_pf) |> 
  View()
```

How about a Parkinsons diagnosis with no features?

```{r}

parkinsonian_data |> 
  dplyr::filter(
    `Parkinsons` == "Yes", !any_pf)
```

No cases; all have features:

```{r}
parkinsonian_data |> 
  dplyr::select(
    `Parkinsons`, any_pf) |> 
  table(useNA = 'always')
```

Can we complete more of the Parkinsons variable?

```{r}

parkinsonian_data |> 
  arrange(`FXS ID`, `Visit Date`) |> 
  filter(.by = `FXS ID`,
         any(incomplete_dx_data(Parkinsons)))

```

(Even if this happened, filling in later observations wouldn't help us, 
since we are only using the first observation.)

Checking "Stiff gait" anyway:

```{r}
parkinsonian_data |> 
  arrange(`FXS ID`, `Visit Date`) |> 
  filter(.by = `FXS ID`,
         any(incomplete_dx_data(`Stiff gait`)))
```

