---
title: "SuStaIn Model Fitting"
authors:
  - name: Alexandra Young
  - name: Douglas Ezra Morrison
    # affiliations:
    #   - ref: ucdphs
    corresponding: true
    email: demorrison@ucdavis.edu
    address: One Shields Ave, Davis, CA 95616
format:
  html:
    echo: true
    toc: true
    number-sections: true
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{OS_example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{r}
#| label: setup
library(fxtas)
library(tidyverse)
library(reticulate)
reticulate::use_condaenv("fxtas39", required = TRUE)
use_rds = FALSE
```

# Import data

```{r}
#| label: import-data

df = sim_data

SuStaInLabels = biomarkers = c('Biomarker 1', 'Biomarker 2', 'Biomarker 3',
               'Biomarker 4', 'Biomarker 5')

biomarker_group_list = list(
  "group 1" = c("Biomarker 1", "Biomarker 2"),
  "group 2" = c("Biomarker 3", "Biomarker 4"),
  "group 3" = "Biomarker 5"
)
```

# Set up data for new version of SuStaIn
The objective here is two create to variables:

* `p_NL` -- a subject x region matrix indicating the probability
that a region is normal for a subject.
* `p_score` -- a subject x (region * score) matrix indicating 
the probability that a region is a given score.

Ps should be non-zero. 

```{r}
#| label: prob-dist
ModelScores = DataScores = as.character(0:3)

control_data = df |> dplyr::filter(Category == "Control") |> dplyr::select(all_of(biomarkers))
patient_data = df |> dplyr::filter(Category == "Patient") |> dplyr::select(all_of(biomarkers))

# prob_correct = (control_data == 0) |> colMeans()
prob_correct = 
  df |> 
  dplyr::filter(Category == "Control") |> 
  summarize(
    across(
      .cols = all_of(biomarkers), 
      .fns = ~ mean(.x == 0))
  ) |> 
  unlist()

prob_dist_dims = 
  list(
    modelscore = ModelScores,
    datascore = DataScores,
    biomarker = biomarkers)

prob_dist = array(
  NA,
  dim = sapply(length, X = prob_dist_dims),
  dimnames = prob_dist_dims)

for (biomarker in biomarkers)
{
  for (datascore in DataScores)
  {
    for (modelscore in ModelScores)
    {
      if(datascore == modelscore)
      {
        prob_dist[modelscore, datascore, biomarker] = prob_correct[biomarker] 
      } else
      {
        prob_dist[modelscore, datascore, biomarker] = 
          (1 - prob_correct[biomarker] ) / (length(DataScores) - 1)
      }
    }
  }
}

prob_nl_dist = prob_dist[1, , ]
prob_score_dist = prob_dist[-1, , ]

```

```{r}
#| label: prob-score
prob_score_dims = 
  list(
  ID = df |> dplyr::filter(Category == "Patient") |> dplyr::pull(Index),
  Biomarker = biomarkers,
  model = ModelScores
)
  
prob_score0 = array(
  NA,
  dim = prob_score_dims |> sapply(length),
  dimnames = prob_score_dims
)

for (biomarker in biomarkers)
{
  for (datascore in DataScores)
  {
    for (modelscore in ModelScores)
    {
      prob_score0[patient_data[[biomarker]] == datascore, biomarker, modelscore] = 
        prob_dist[modelscore, datascore, biomarker]
    }
    
  }
}

prob_nl = prob_score0[,,"0"]
prob_score = prob_score0[,,-1]

```

```{r }
#| label: build-score-vals-only-observed

biomarker_levels = 
  df |> 
  dplyr::select(all_of(biomarkers)) |> 
  lapply(FUN = \(x) x |> unique() |> sort())
score_vals = build_score_vals(biomarker_levels)

score_vals

```

or, to match original example from Alex Young:

```{R}
#| label: build-score-vals-all-possible

biomarker_levels = list(
  "Biomarker 1" = 0:3,
  "Biomarker 2" = 0:3,
  "Biomarker 3" = 0:3,
  "Biomarker 4" = 0:3,
  "Biomarker 5" = 0:3
)

score_vals = build_score_vals(biomarker_levels)

score_vals
```

::: {.callout-warning}
When switching between these two methods of calculating `score_vals`,
the number of events in the sequence changes, so the pickle files need 
to be rebuilt accordingly. 
Otherwise, we might get errors about array lengths and indices.
:::

# Run SuStaIn model

```{r "set run parameters"}

N_startpoints = 10L
N_S_max = 4L
N_iterations_MCMC = 1e5L
dataset_name = 'sample_data'
output_folder = 
  fs::path("inst/extdata/sim_data")
print(output_folder)
```

```{python "initialize-model"}

import pandas
import numpy as np
import numpy.matlib
import matplotlib.pyplot as plt
import seaborn as sns
import sys
import scipy
from glob import glob
import pySuStaIn
import os

os.getcwd()
print(r.output_folder)
sustain_input = pySuStaIn.OrdinalSustain(
    r.prob_nl,
    r.prob_score,
    r.score_vals,
    r.SuStaInLabels,
    r.N_startpoints,
    r.N_S_max, 
    r.N_iterations_MCMC, 
    r.output_folder, 
    r.dataset_name, 
    True,
    1)
```

```{python "fit-model", fig.width = 8, fig.height = 9}

[samples_sequence, samples_f, 
ml_subtype, prob_ml_subtype, ml_stage, 
prob_ml_stage, prob_subtype_stage, 
samples_likelihood] = sustain_input.run_sustain_algorithm(plot=True)

```

In R, we can run the algorithm with `run_OSA()`:

```{r}
#| label: run-osa-R

biomarker_group_list = list(
  "group 1" = biomarkers[1:2],
  "group 2" = biomarkers[3:4],
  "group 3" = biomarkers[5]
)

biomarker_groups = df |> compile_biomarker_groups_table(
  biomarker_group_list = biomarker_group_list
)
patient_data     |> readr::write_rds(file = fs::path(output_folder, "data.rds"))
biomarker_levels |> readr::write_rds(file = fs::path(output_folder, "biomarker_levels.rds"))
biomarker_groups |> readr::write_rds(file = fs::path(output_folder, "biomarker_groups.rds"))
  
osa_output = run_and_save_OSA(
  dataset_name = dataset_name,
  patient_data = patient_data,
  prob_correct = prob_correct,
  prob_score = prob_score0,
  score_vals = score_vals,
  output_folder = output_folder, 
  biomarker_levels = biomarker_levels,
  SuStaInLabels = SuStaInLabels,
  N_startpoints = N_startpoints,
  N_iterations_MCMC = N_iterations_MCMC,
  plot = TRUE,
  N_S_max = N_S_max,
  N_CV_folds = 10,
  rerun = TRUE,
  verbose = TRUE
)

cvic_sim_data <- osa_output |> extract_cvic()
cvic_sim_data |> usethis::use_data(overwrite = TRUE)
```

::: {#fig-cvic layout-ncol="2"}

::: {#fig-cvic-1}

```{r}
cvic_sim_data$CVIC |> plot_CVIC()
```

:::

::: {#fig-cvic-2}

```{r}
cvic_sim_data$loglike_matrix |> plot_cv_loglik()
```

:::

**Cross-validation information criterion (CVIC)**

:::

```{r}
#| label: graph-likelihoods
likelihoods = 
  py$samples_likelihood |> 
  magrittr::set_colnames(
    paste(1:ncol(py$samples_likelihood), "subtype(s)")
  ) |> 
  as_tibble() |> 
  dplyr::mutate(Iteration = dplyr::row_number())

alpha = 0.5
ggplot() +
  geom_line(
    alpha = alpha,
    aes(
    x = seq_len(nrow(likelihoods)), 
    y = likelihoods[[1]],
    col = "One subtype"
    )
  ) +
  geom_line(
    alpha = alpha,
    aes(
    x = seq_len(nrow(likelihoods)), 
    y = likelihoods[[2]],
    col = "Two subtypes"
    )
  ) +
  geom_line(
    alpha = alpha,
    aes(
    x = seq_len(nrow(likelihoods)), 
    y = likelihoods[[3]],
    col = "Three subtypes"
    )
  )

```

```{r}
#| label: graph-likelihoods-2
likelihoods2 = 
  likelihoods  |> 
  tidyr::pivot_longer(cols = paste(1:ncol(py$samples_likelihood), "subtype(s)"))

likelihoods2 |> 
  ggplot(
       aes(
    x = Iteration, 
    y = value,
    col = name
    )) +
  geom_line(alpha = alpha) +
  facet_wrap(~name) +
  ylab("log-likelihood")

```

### trying `plt.show()`

```{python}
#| label: fig-python.plt.show
#| fig-cap: "python.plt.show"
# plt.show() # doesn't work
```


### attempting to call the plot function to get the heatmaps

```{python}
#| label: fig-plot_positional_var
#| fig-cap: "calling `plot_positional_var()` directly in python"
n_samples = r.prob_nl.shape[0]
figs, ax = sustain_input.plot_positional_var(
  samples_sequence=samples_sequence,
  samples_f=samples_f,
  n_samples= n_samples,
  score_vals = r.score_vals,
  biomarker_labels=r.biomarkers,
  save_path=r.output_folder)
```

show figs:

```{python}
#| label: fig-show-figs
for fig in figs:
  fig.show()

# py$samples_sequence

```

## figs in R

```{r}
#| label: fig-plot-compact-pvd-n1
#| fig-cap: "Compact PVD: 1 subgroup"
#| fig-height: 3
figs = extract_figs_from_pickle(
  output_folder = output_folder,
  use_rds = use_rds,
  n_s = 1)
  
figs = list(figs)

plot_compact_pvd(
  figs = figs,
  tile_height = 1,
  facet_label_prefix = "All data",
  y_text_size = 8,
  biomarker_order = biomarkers,
  scale_colors = c("red", "magenta", "blue")
)
```

```{r}
#| label: fig-plot-compact-pvd-n2
#| fig-cap: "Compact PVD: 2 subgroups"
#| fig-height: 3

figs = extract_figs_from_pickle(
  output_folder = output_folder,
  use_rds = use_rds,
  n_s = 2)
  
plot_compact_pvd(
  figs = figs,
  tile_height = 1,
  y_text_size = 8,
  biomarker_order = biomarkers,
  scale_colors = c("red", "magenta", "blue")
)
```

```{r}
#| label: fig-plot-compact-pvd-n3
#| fig-cap: "Compact PVD: 3 subgroups"
#| fig-height: 3

results3 <- extract_results_from_pickle(
  output_folder = output_folder,
  use_rds = use_rds,
  n_s = 3)

results_s3_sim_data <- results3
results_s3_sim_data |> use_data(overwrite = TRUE)

figs3 = extract_figs_from_pickle(
  output_folder = output_folder,
  use_rds = use_rds,
  n_s = 3)

plot_compact_pvd(
  figs = figs3,
  tile_height = 1,
  facet_label_prefix = NULL,
  y_text_size = 8,
  biomarker_order = biomarkers,
  scale_colors = c("red", "magenta", "blue")
)
```

we can also do position difference graphs: (PDGs):

```{r}
figs[1:2] |> pvd_lineplot()
```

We can check the distribution of ML stages with `stage_barplot()`:

```{r}
results3 |> stage_barplot()
```


```{r}
#| label: save-files

readr::write_rds(
  x = biomarker_levels,
  file = fs::path(output_folder, "biomarker_levels.rds")
)

biomarker_groups_table = df |> compile_biomarker_groups_table(
 biomarker_group_list
)

readr::write_rds(
  x = biomarker_groups_table,
  file = fs::path(output_folder, "biomarker_groups.rds")
)

```


